---
title: "Song Lyrics Data Story"
author: "Sagar Lal"
date: "9/14/2019"
output:
  html_document: default
  pdf_document: default
---

"lyrics_filter.csv" is a filtered corpus of 380,000+ song lyrics from from MetroLyrics. You can read more about it on [Kaggle](https://www.kaggle.com/gyani95/380000-lyrics-from-metrolyrics).

"info_artist.csv" provides the background information of all the artistis. These information are scraped from [LyricsFreak](https://www.lyricsfreak.com/).

Here, we explore these data sets and try to find interesting patterns.

```{r load libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(tidytext)
library(plotly)
library(DT)
library(tm)
library(data.table)
library(scales)
library(wordcloud2)
library(gridExtra)
library(ngram)
library(shiny) 
library(ggplot2)
library(ggridges)
```


#### This notebook was prepared with the following environmental settings.

```{r version, echo=FALSE}
#This notebook was prepared with the following environmental settings.
print(R.version)
```

```{r load data, warning=FALSE, message=FALSE, echo=FALSE}
#We use the processed data and artist information for our analysis.

# load lyrics data
load('../output/processed_lyrics.RData') 
# load artist information
dt_artist <- fread('../data/artists.csv') 
```

# **Goal: Explore how genre's change over time in our dataset.**


#### Each genre's song output

```{r songs per genre per year, echo=FALSE}
#Find number of songs per genre in a given year
genre_year_song_counts = dt_lyrics %>% group_by(genre,year) %>% mutate(genre_year_count = n()) %>% ungroup() %>% select(genre, year, genre_year_count) %>% unique() %>% filter(genre != "Other" & genre != "Not Available" & year >2003) 

ggplot(data = genre_year_song_counts, 
       mapping = aes(x = year, y = genre_year_count, color = genre)) +
  geom_line() +
  facet_grid(rows = vars(genre), scales = "free_y") +
  xlab("Year") +
  ylab(label="Number of Songs") +
  ggtitle("Number of songs in each genre in a year") + 
  theme_bw()
```

Most genre's appear to have their maximal output of songs in 2006 with rock achieveing a peak output of 25k songs. A few genre's have a peak output in 2007. This includes country, indie, and metal. It is notable that almost all genres produce significantly more songs between 2005-2008 before receding to consistently fewer songs, suggesting the dataset aggregated music from a specific time. Note the y-axis is different across plots, due to certain genres such as Indie and R&B producing significantly less music. 


#### How pervasive are lyrics in different genres


```{r trigrams popularity in genre over time, echo=FALSE}
#Only care about trigram appearing once in a song
lyric_trigrams <- dt_lyrics %>%
      unnest_tokens(trigram, stemmedwords, token = "ngrams", n = 3) %>% unique()

#Need to seperate words so can determine that words aren't the same
trigram_seperated <- lyric_trigrams %>%
      separate(trigram, c("word1", "word2", "word3"), sep = " ", remove = FALSE)

#Make sure trigrams don't have same word to avoid "baby baby baby", also remove case of each song written by babyface, also only care about phrase appearing once in a song (don't want 'i am blue' to repeatedly appear)
unique_trigrams = trigram_seperated %>%
                    filter(word1 != word2 & word2 != word3 & word1 != word3)

#Find how many times each trigram appears in songs of a particular genre
trigram_genre_counts = unique_trigrams %>% group_by(genre, trigram) %>% mutate(count = n()) %>% ungroup()

#Filter out specific trigrams that aren't lyrics, as well as unknown and not avaialable genre
trigram_genre_counts = trigram_genre_counts %>% filter(
                            genre != "Not Available" &
                            genre != "Other" &
                            trigram != "written reid babyface" &
                            trigram != "written babyface performed" &
                            trigram != "chorus repeat x" &
                            trigram != "daryl simmons performed" &
                            trigram != "written billy burnette" &
                            trigram != "babyface daryl simmons" &
                            trigram != "dennis ufo music" &
                            trigram != "sandy dennis ufo" &
                            trigram != "reid babyface daryl" &
                            trigram != "performed tony braxton" &
                            trigram != "music vestergaard lyrics"&
                            trigram != "vestergaard lyrics frolund" &
                            trigram != "written babyface reid" &
                            trigram != "performed whitney houston")


#Find most frequent appearing trigram in each genre and the number of times
genre_top_percentile = trigram_genre_counts %>% group_by(genre) %>% summarise(most_popular = quantile(count, probs = 1))

#Rejoin so only keep songs that contain most popular trigram in that song's genre
most_popular_trigram = trigram_genre_counts %>% inner_join(genre_top_percentile, by = c("genre" = "genre", "count" = "most_popular"))

#Find frequency that that trigram for songs in a genre appeared in each year
trigram_over_time = most_popular_trigram %>% group_by(genre, trigram, year) %>% mutate(genre_popular_trigram_count_by_year = n()) %>% ungroup()

trigram_over_time = trigram_over_time %>% left_join(genre_year_song_counts, by = c("genre" = "genre", "year" = "year")) %>% mutate(percent_of_songs_in = genre_popular_trigram_count_by_year/genre_year_count * 100)

#Create unique column for graph
trigram_over_time$genre_w1_w2_w3 = paste(trigram_over_time$genre, trigram_over_time$trigram, sep="_")


#Select columns for graph and remove ties
trigram_genre_year_data = trigram_over_time %>% 
  select(year, percent_of_songs_in, genre_w1_w2_w3, count) %>% 
  unique() %>% 
  filter(
      genre_w1_w2_w3 != "Indie_time fall love" & 
      genre_w1_w2_w3 != "Jazz_sing love song" & 
      genre_w1_w2_w3 != "Metal_blaze ya dead" & 
      genre_w1_w2_w3 != "R&B_love rest life")

#To improve graph remove rows where data only shows up once
trigram_genre_year_data = trigram_genre_year_data %>% filter(year > 2000)

trigram_genre_year_data[is.na(trigram_genre_year_data)] <- 0

ggplot(data = trigram_genre_year_data, 
       mapping = aes(x = year, y = percent_of_songs_in, color = genre_w1_w2_w3)) +
  geom_line() +
  facet_grid(rows = vars(genre_w1_w2_w3), scales = "free_y") +
  xlab("Year") +
  ylab(label="% of songs in during year") +
  ggtitle("Each Genre's most popular trigram appearances over time") + 
  theme_bw()


```

For each genre, the trigram that appeared in the most number of songs across the dataset was computed. For each year, the percentage of the genre's songs that the most-popular trigram appeared in was computed. It is notable that most of the popular phrases appear early on and then gain popularity over time before having a dip usually between 2013-2016. The exceptions to this are hip-hop and metal, which seem to be peaking. This is potentially indicative of lyrical similarities between mid 2000s and current lyrics or sentiments. Metal seems to be an outlier case where one phrase was incredibly popular in 2007-2008, though it also corresponds to a decline in total nubmer of songs at the time.


#### Genre sentiment over time


``` {r Genre change in sentiment over time, echo=FALSE}
#Create tidy data where row contains one word in a song
unnested_lyrics <- dt_lyrics %>%
      unnest_tokens(word, lyrics)

#Replace positive and negative with 1, -1
new_sentiments <- sentiments %>% 
  mutate( sentiment = ifelse(sentiment == "positive", 1, ifelse(sentiment == "negative", -1, 0))) 

#Join with lyrics so can have a measurement of songs sentiment
song_sentiment <- unnested_lyrics %>% left_join(new_sentiments, by = c("word" = "word"))
song_sentiment[is.na(song_sentiment)] <- 0

#Compute average song sentiment which is sum of sentiment of words in song/total words in song
average_song_sentiment <- song_sentiment %>%
  group_by(song,genre,artist,year) %>% summarize(total_song_sentiment = sum(sentiment), song_lyric_words = n()) %>%   mutate(average_song_sentiment = total_song_sentiment/song_lyric_words) %>% ungroup()

#Compute average genre sentiment by taking sum of average sentiment of songs in genre in year/total number of songs in genre in that year
average_genre_sentiment_over_time <- average_song_sentiment %>% group_by(genre,year) %>% summarize(total_genre_sentiment = sum(average_song_sentiment), total_genre_songs = n()) %>%
  mutate(average_genre_sentiment = total_genre_sentiment/total_genre_songs) %>% ungroup() %>% filter(year>1999 & genre != "Other" & genre != "Not Available")

ggplot(data = average_genre_sentiment_over_time, 
       mapping = aes(x = year, y = average_genre_sentiment, color = genre)) +
  geom_line() +
  facet_grid(rows = vars(genre)) +
  xlab("Year") +
  ylab(label="Average Sentiment per Song") +
  ggtitle("Genre Sentiment over Time") + 
  theme_bw()
```


Most genre's have a fairly stable average sentiment per song over time, which is based on the individual sentiment of words. The exception to this is electronic music, which is pretty consistently dropping in sentiment over time. While metal is consistently the most negative genre, there are a few that tend to be more positive (Country, Jazz, Pop, and R&B). Given that averages can be skewed by outliers, we next want to examine the distribution over time

``` {r joy plots, echo=FALSE}

#CREATE JOY PLOT AND POINT OUT SPECIFIC EXAMPLE

ridge_plot = average_song_sentiment %>% filter(year>1999 & genre != "Other" & genre != "Not Available")

ggplot(ridge_plot, aes(y=as.factor(year),
                      x=average_song_sentiment,
                      fill=genre)) +
  geom_density_ridges(alpha=0.25) +
  scale_y_discrete(expand = c(0.01, 0)) +  
  scale_x_continuous(expand = c(-.5, .5))

ggplot(ridge_plot, mapping = aes(y=as.factor(year),
                      x=average_song_sentiment)) +
  geom_density_ridges(alpha=0.25) +
    facet_grid(rows = vars(genre)) +

  scale_y_discrete(expand = c(0.01, 0)) +  
  scale_x_continuous(expand = c(-.5, .5))

```


